---
title: "Normalize expression data from Human Cell Landscape (HCL)"
author: "Shuyang Yao, shuyang.yao@ki.se"
date: "2020-06-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(data.table)
library(ggplot2)
library(org.Hs.eg.db)
```
  
Data from: https://figshare.com/articles/HCL_DGE_Data/7235471  
Paper: PMID 32214235  
  
Calculated specificity according to PMID: 32341526  
  
Steps:  
1. Check files in expression data folder and annotation folder.  
Note: expression data: row=gene, col=cellID; annotation data: row=cellID, col=cell type (among other info, e.g., batch, sample...)  
2. Get expression and scaled expression for each gene per organ_cellType; combine btw organ_cellTypes.  
3. Calculate specificity  
4. add ENTREZ and ensembl ID  
  
Step 1. Check files  
```{r}
# 1.1 files in expression folder
mypath <- "/nas/depts/007/sullilab/shared/sy/HCL/00data/" # longleaf
a <- paste(mypath, "dge_rmbatch_data", sep="")
b <- list.files(a, full.names = T) %>% as.tibble() %>% 
  mutate(filename=basename(value))
b$tissuename <- gsub(pattern = "_rmbatchdge.txt.gz",replacement = "",x = b$filename)

# 1.2 file in annotation folder
c <- paste(mypath, "annotation_rmbatch_data", sep="")
d <- list.files(c, full.names = T) %>% as.tibble() %>% 
  mutate(filename=basename(value))
d$tissuename <- gsub(pattern = "_rmbatchAnno.csv",replacement = "",x = d$filename)

head(b)
head(d)

# check batch
for(i in 1:nrow(d)){
  tmp <- fread(d$value[i], stringsAsFactors = F, data.table = F) %>% distinct(Batch)
  if(nrow(tmp)>1){print(paste(i,". Tissue ", d$tissuename[i], " has ",nrow(tmp)," batches.", sep=""))}
}
print("Each of the other tissues has one batch.")

```
  
*Check all tissue names and organ_celltype names.*  
```{r}
rm(organlist)
organlist <- list()
for (i in 1:length(d$tissuename)){
  tmp<- fread(d$value[i], stringsAsFactors = F, data.table = F) %>% 
    mutate(organ_cellType=paste(Sample, Celltype, sep="_")) %>%
    distinct(Sample, organ_cellType) %>%
    dplyr::rename(organ=Sample)
  tmp$organ_cellType <- gsub(" ","_",tmp$organ_cellType)
  tmp$organ_cellType <- gsub("/","_or_",tmp$organ_cellType)
  tmp$organ_cellType <- gsub("<a0>","_",tmp$organ_cellType)
  tmp$organ_cellType <- gsub("<fc>","u",tmp$organ_cellType)
  tmp$organ_cellType <- gsub("\\(","_",tmp$organ_cellType)
  tmp$organ_cellType <- gsub("\\)","_",tmp$organ_cellType)
  tmp$organ_cellType <- gsub("\n","",tmp$organ_cellType) 
  tmp$organ_cellType <- gsub("__","_",tmp$organ_cellType)
  tmp$organ_cellType <- gsub("__","_",tmp$organ_cellType)
  tmp$organ_cellType <- gsub("_$","",tmp$organ_cellType)
  organlist[[i]] <- tmp
}
rm(tmp)
tmp <- rbindlist(organlist)

# check all organs:
organ_check <- tmp %>% distinct(organ) 
#= found one typo: change "FetalIntetsine" to "FetalIntestine" in the following processing

# check organ_cellType
tmp <- tmp %>% dplyr::arrange(organ, organ_cellType)
tmp$organ_corrected <- gsub("FetalIntetsine","FetalIntestine",tmp$organ)
tmp$organ_cellType_corrected <- gsub("FetalIntetsine","FetalIntestine",tmp$organ_cellType)
tmp <- tmp %>% dplyr::count(organ_corrected, organ_cellType_corrected) %>% dplyr::filter(n>1) # 201 organ_cellTypes that were duplicated 2-4 times.
fwrite(tmp,
       paste(mypath,"Duplicated_organ_cellType.txt",sep=""), 
       sep="\t", col.names = T)

```
  
Step 2. Get expression and scaled expression for each gene per organ_cellType.  
Inside the loop, do this for each orga_cellType:  
1. Read in exp.data; change to long-format: gene, cellID, expression (count);  
2. Merge to cellType (inner_join, rm cells that were not clustered to a cell type), merge by cellID, get: geneName, cellID, expression, cellType; add organ name to cellType: organ_cellType;  
3. Collapse by gene and cellType: geneName, organ_cellType, geneCount;  
4. Combine btw tissues/organs, in long format  
5. After combination, Redo count of genes per organ_cellType (due to duplication of genes and organ_cellType).  
6. Scale per organ_cellType expression (to 10k).  
  
```{r}
rm(i, Final_expdat_celltype_scaled10k_long,result)
result <- list()
for(i in 1:length(b$tissuename)){
  print(b$tissuename[i])
  print(d$tissuename[i])
  if(d$tissuename[i]==b$tissuename[i]){
    # 1. exp.data: change to long-format: gene, cellID, expression
    expdat <- fread(b$value[i], stringsAsFactors = F, data.table = F) %>%
      dplyr::rename(geneName=V1)
    expdat[is.na(expdat)] <- 0
    expdat_long <- expdat %>% gather(cellID, count, -geneName)
    rm(expdat)
    # 2. merge to cellType, merge by cellID, get: geneName, cellID, expression, cellType; add organ name to cellType: organ_cellType.
    annotdat <- fread(d$value[i], stringsAsFactors = F, data.table = F) %>%
      dplyr::rename(Sample_old=Sample) %>% 
      mutate(Sample=ifelse(Sample_old=="FetalIntetsine", "FetalIntestine", Sample_old)) 
    annotdat <- annotdat %>%
      mutate(organ_cellType=paste(Sample, Celltype, sep="_")) %>%
      dplyr::rename(cellID=Cell_id) %>%
      dplyr::select(cellID, organ_cellType)
    annotdat$organ_cellType <- gsub(" ","_",annotdat$organ_cellType)
    annotdat$organ_cellType <- gsub(pattern = "\n",replacement = "",annotdat$organ_cellType)
    annotdat$organ_cellType <- gsub("/","_or_",annotdat$organ_cellType)
    annotdat$organ_cellType <- gsub("<a0>","_",annotdat$organ_cellType)
    annotdat$organ_cellType <- gsub("<fc>","u",annotdat$organ_cellType)
    annotdat$organ_cellType <- gsub("\\(","_",annotdat$organ_cellType)
    annotdat$organ_cellType <- gsub("\\)","_",annotdat$organ_cellType)
    annotdat$organ_cellType <- gsub("__","_",annotdat$organ_cellType)
    annotdat$organ_cellType <- gsub("__","_",annotdat$organ_cellType)
    annotdat$organ_cellType <- gsub("__","_",annotdat$organ_cellType)
    annotdat$organ_cellType <- gsub("_$","",annotdat$organ_cellType) # remove the "_" if it is the last character
    expdat_long <- expdat_long %>% inner_join(annotdat, by="cellID")
    expdat_celltype_long <- expdat_long %>% 
      group_by(geneName, organ_cellType) %>% 
      summarise(geneCount=as.numeric(sum(count))) %>% 
      as.data.frame() %>% 
      dplyr::select(geneName, organ_cellType, geneCount)
    
    result[[i]] <- expdat_celltype_long
  }else{
    stop("Tissue name did not match in exp.data and annot.data.")
  }
}

# 4. Combine btw files
Final_expdat_celltype_scaled10k_long <- rbindlist(result) 

# 5. To fix issue with dupliated organ_cellType: redo geneCount, 
# so each pair of geneName and organ_cellType appears only once.
Final_expdat_celltype_scaled10k_long <- Final_expdat_celltype_scaled10k_long %>%
  group_by(geneName, organ_cellType) %>% 
  summarise(geneCount2=as.numeric(sum(geneCount))) %>%
  as.data.frame() %>% 
  dplyr::select(geneName, organ_cellType, geneCount2) %>%
  dplyr::rename(geneCount=geneCount2)

# 6. scaling after combination
Final_expdat_celltype_scaled10k_long <- Final_expdat_celltype_scaled10k_long %>%
  group_by(organ_cellType) %>% 
  mutate(geneCount_scaled10k=geneCount*10000/sum(geneCount)) %>% 
  as.data.frame()

```
  
Step 3. Calculate specificity   
1. Remove genes that were not expressed in any tissue  
2. Calculate specificity  

```{r}
# 1. Remove genes not expressed in any tissues  
g2rm <- Final_expdat_celltype_scaled10k_long %>% 
  group_by(geneName) %>% 
  summarise(sum_exp=sum(geneCount)) %>% 
  filter(sum_exp==0)
Final_expdat_celltype_scaled10k_long <- filter(Final_expdat_celltype_scaled10k_long, !geneName%in%g2rm$geneName)

# 2. Calculate specificity
Final_expdat_celltype_scaled10k_long <- Final_expdat_celltype_scaled10k_long %>% 
  group_by(geneName) %>% 
  mutate(spe_10k=geneCount_scaled10k/sum(geneCount_scaled10k)) %>%
  ungroup()
```
  
Step 4. Add ENTREZ ID  
  
```{r}
entrez_ensembl <- AnnotationDbi::toTable(org.Hs.eg.db::org.Hs.egENSEMBL)
entrez_symbol <- AnnotationDbi::toTable(org.Hs.eg.db::org.Hs.egSYMBOL)
entrez_ensembl <- inner_join(entrez_ensembl,entrez_symbol, by = "gene_id")
```
  
Only keep genes with a unique entrez and ensembl id.  
  
```{r,message=FALSE}
entrez_ensembl_unique_genes_entrez <- entrez_ensembl %>% count(gene_id) %>% filter(n==1)
entrez_ensembl_unique_genes_ens <- entrez_ensembl %>% count(ensembl_id) %>% filter(n==1)
entrez_ensembl_unique_genes_symbol <- entrez_ensembl %>% count(symbol) %>% filter(n==1)

entrez_ensembl <- entrez_ensembl %>% filter(gene_id%in%entrez_ensembl_unique_genes_entrez$gene_id & 
                                              ensembl_id %in% entrez_ensembl_unique_genes_ens$ensembl_id & 
                                              symbol %in% entrez_ensembl_unique_genes_symbol$symbol)
colnames(entrez_ensembl) <- c("ENTREZ","gene_id","geneName") # original names: c("gene_id","ensembl_id","symbol")
```
  
Merge to long-format matrix  
```{r}
Final_expdat_celltype_scaled10k_long <- left_join(Final_expdat_celltype_scaled10k_long,entrez_ensembl,by="geneName") %>% 
  dplyr::rename("Gene"="geneName") %>%
  as.data.frame() 

myfile <- paste(mypath,"Final_expdat_celltype_scaled10k_long.txt.gz",sep="")
Final_expdat_celltype_scaled10k_long %>% write_tsv(gzfile(myfile))

```
  
Save expression profile for other processing  
```{r, eval=F}
save(Final_expdat_celltype_scaled10k_long,
     file = paste(mypath, "expression.ready.Rdata", sep=""))
```
  
Step 5. Write MAGMA and LDSC output  
Get coordinate  
data: NCBI37.3.gene.loc.extendedMHCexcluded from repository: https://github.com/jbryois/scRNA_disease  
```{r}
gene_coordinates <- 
  read_tsv(paste(mypath,"/Other/NCBI37.3.gene.loc.extendedMHCexcluded",sep=""), col_names = FALSE,col_types = 'cciicc') %>%
  mutate(start=ifelse(X3-100000<0,0,X3-100000),end=X4+100000) %>%
  dplyr::select(X2,start,end,1) %>% 
  dplyr::rename(chr=X2, ENTREZ=X1) %>% 
  mutate(chr=paste0("chr",chr))

# keep only genes with coordinates
exp_dat <- Final_expdat_celltype_scaled10k_long %>% 
  filter(ENTREZ %in% gene_coordinates$ENTREZ)
```
  
Number of genes to keep (in top 10%)  
```{r}
n_genes_to_keep <- length(unique(exp_dat$ENTREZ))*0.1
n_genes_to_keep <- round(n_genes_to_keep)
```
  
MAGMA  
```{r}
magma_top10 <- function(d,organ_cellType){
  d_spe <- d %>% filter(is.na(ENTREZ)==F) %>% 
    group_by(organ_cellType) %>% 
    filter(geneCount_scaled10k > 0.01) %>%
    top_n(., n_genes_to_keep, spe_10k) 
  d_spe %>% do(write_group_magma(.,organ_cellType))
}

write_group_magma  = function(df,organ_cellType) {
  df <- dplyr::select(df,organ_cellType,ENTREZ)
  df_name <- make.names(unique(df[1]))
  colnames(df)[2] <- df_name  
  dir.create(paste0("MAGMA/"), showWarnings = FALSE)
  dplyr::select(df,2) %>% t() %>% as.data.frame() %>% rownames_to_column("Cat") %>%
    write_tsv("MAGMA/top10.txt",append=T)
  return(df)
}

# write
setwd(mypath)
exp_dat %>% magma_top10("HCL")
```
  
LDSC  
```{r}
genesets <- exp_dat %>% filter(is.na(ENTREZ)==F) %>% 
  group_by(organ_cellType) %>% 
  filter(geneCount_scaled10k > 0.01) %>%
  top_n(., n_genes_to_keep, spe_10k) %>%
  ungroup()

# output directory
outdir <- paste(mypath,"LDSC", sep="")
# loop over tissues: merge for position and output
tissues <- genesets %>% distinct(organ_cellType)
rm(i,dataseti,tissuei)
mylist <- list()
for (i in 1:nrow(tissues)){
  tissuei <- tissues$organ_cellType[i]
  dataseti <- genesets %>% filter(organ_cellType %in% !!tissuei) %>%
    mutate(ENTREZ=as.character(ENTREZ)) %>%
    dplyr::select(ENTREZ, organ_cellType)
  dataseti <- dataseti %>% left_join(gene_coordinates, by=c("ENTREZ")) %>%
    dplyr::select(chr, start, end) %>%
    filter(is.na(start)==F)
  tissueiname <- gsub("/","_or_",tissuei)
  tissueiname <- gsub("<a0>","_",tissuei)
  tissueiname <- gsub("<fc>","u",tissuei)
  tissueiname <- gsub(">","_",tissuei)
  tissueiname <- gsub("\\(","_",tissueiname)
  tissueiname <- gsub("\\)","_",tissueiname)
  tissueiname <- gsub("__","_",tissueiname)
  tissueiname <- gsub("__","_",tissueiname)
  tissueiname <- gsub("__","_",tissueiname)
  outnamei <- paste(outdir,"/",tissueiname,".bed", sep="")
  fwrite(dataseti, outnamei, sep="\t", col.names = F)
}

```
  
  