---
title: "Enrichment analysis results"
author: "Shuyang Yao (shuyang.yao@ki.se)"
date: "1/8/2021"
output: 
  html_document:
    toc: true
    toc_float: true
    fig_caption: yes
    theme: sandstone
    highlight: haddock
---
<style type="text/css">
/* Whole document: */
body{
  font-family: Helvetica;
  font-size: 11pt;
}
/* Headers */
h1{
  font-size: 18pt;
}
h2{
  font-size: 16pt;
}
h3{
  font-size: 14pt;
}
h4{
  font-size: 12pt;
}
</style>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Enrichment analysis results for MDD3, in expression datasets: GTEx (version 8), Zeisel, Zeisel_lvl5, and HCL  
Output:  
- Data frames, separately for each expression dataset  
- Graphs, separately for each expression dataset: x=mean($-log_{10} P$), y=tissue/cell types  
  
```{r}
library(data.table)
library(tidyverse)
library(ggplot2)
```
  
Functions:  
1. Output table  
```{r}
myfun.data <- function(inpath_ldsc, inpath_magma, trait, outname_df){
e1 <- fread(inpath_ldsc, stringsAsFactors = F, data.table = F) %>%
  mutate(Method="LDSC")
e1$FDR <- p.adjust(e1$P, method="fdr", n=nrow(e1) )

e2 <- fread(inpath_magma, stringsAsFactors = F, data.table = F) 
if("FULL_NAME" %in% colnames(e2)){
  e2 <- e2 %>%
  mutate(Trait=!!trait, Method="MAGMA") %>%
  rename(Tissue=FULL_NAME, Coefficient=BETA, Coefficient_std_error=SE) %>%
  select(Trait, Tissue, P, Coefficient, Coefficient_std_error, Method)
}else{
  e2 <- e2 %>%
  mutate(Trait=!!trait, Method="MAGMA") %>%
  rename(Tissue=VARIABLE, Coefficient=BETA, Coefficient_std_error=SE) %>%
  select(Trait, Tissue, P, Coefficient, Coefficient_std_error, Method)
}
e2$FDR <- p.adjust(e2$P, method="fdr", n=nrow(e2) )

if(expdataset=="HCL"){
  # special for HCL: fix names:
  e1$Tissue <- gsub("-","_", e1$Tissue)
  e1$Tissue <- gsub("\\+","", e1$Tissue)
  e2$Tissue <- gsub("\\.","_", e2$Tissue)
  e2$Tissue <- gsub("__","_", e2$Tissue)
  e2$Tissue <- gsub("AdultMuscle_Striated_muscle_a0_cell","AdultMuscle_Striated_muscle_cell", e2$Tissue)
  e2$Tissue <- gsub("FetalEyes_M_fc_ller_glia","FetalEyes_Muller_glia", e2$Tissue)
  e2$Tissue <- gsub("FetalMuscle_Autonomic_nervous_system_a0_","FetalMuscle_Autonomic_nervous_system", e2$Tissue)
}

# combine to get dataset
e3 <- rbind(e1,e2) %>%
  mutate(minuslog10P=-log10(P)) %>%
  group_by(Tissue) %>%
  summarise(minuslog10P_mean=mean(minuslog10P)) %>%
  ungroup()
e <- rbind(e1,e2) %>% left_join(e3, by="Tissue")
if(expdataset=="HCL"){
  e <- e %>% filter(!grepl("Unknown", Tissue))
}

# write out
fwrite(e, outname_df, sep="\t", col.names = T)
return(e)
}

```
  
2. Plot  
```{r}
myfun.plot <- function(e, expdataset){
  pthresh <- 0.05/(nrow(e)/2)
 
f1 <- e %>% filter(Method=="LDSC") %>% mutate(sig_ldsc=ifelse((P<=pthresh), 1, 0)) %>% select(Tissue, sig_ldsc)
f2 <- e %>% filter(Method=="MAGMA") %>% mutate(sig_magma=ifelse((P<=pthresh), 1, 0)) %>% select(Tissue, sig_magma)
#f <- full_join(f1, f2, by="Tissue") %>%
f <- inner_join(f1, f2, by="Tissue") %>%
  mutate(Significant=case_when((sig_ldsc==1 & sig_magma==1) ~ "Both",
                               (sig_ldsc==1 & sig_magma==0) ~ "LDSC",
                               (sig_ldsc==0 & sig_magma==1) ~ "MAGMA",
                               (sig_ldsc==0 & sig_magma==0) ~ "None")) %>%
  left_join(e, by="Tissue") %>%
  distinct(Trait, Tissue, Significant, minuslog10P_mean)

p <- ggplot(f,aes(Tissue,minuslog10P_mean,fill=Significant)) + 
  geom_bar(stat="identity") + coord_flip() + facet_wrap(~Trait) 
p <- p + theme_bw() + geom_hline(yintercept =-log10(0.05/nrow(f))) + xlab("")
p <- p + ylab(expression('mean -log'[10]*'(pvalue)'))
# color:
#cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
if(length(unique(f$Significant))==4){
  mycolor=c("#E69F00", "#56B4E9","#0072B2","#999999")
}
if(length(unique(f$Significant))==3){
  mycolor=c("#E69F00", "#0072B2","#999999")
}
if(length(unique(f$Significant))==2){
  mycolor=c("#E69F00", "#999999")
}
if(length(unique(f$Significant))==1){
  mycolor=c("#999999")
}
p <- p + scale_fill_manual(values=mycolor)
# xlab
p_xlab <- paste("Cell types in ", expdataset, " dataset", sep="")
p <- p + xlab(p_xlab)
mylist <- list(f, p)
return(mylist)
}

```
  
```{r}
plot.magma.ldsc <- function(inpath_magma, inpath_ldsc, plot_out, dataset){
  
  df.magma <- fread(inpath_magma,
                    stringsAsFactors = F, data.table = F) %>%
    mutate(if.sig=ifelse(P<=0.05/nrow(df.magma), 1, 0),
           test="MAGMA") %>%
    select(VARIABLE, BETA, SE, P, if.sig, test)
  tmp <- df.magma %>% arrange(-P) %>% 
    mutate(order=row_number()) %>%
    select(VARIABLE, order)
  df.ldsc <- fread(inpath_ldsc,
                    stringsAsFactors = F, data.table = F)%>%
    rename(VARIABLE=Tissue, BETA=Coefficient, SE=Coefficient_std_error) %>%
    mutate(if.sig=ifelse(P<=0.05/nrow(df.magma), 1, 0),
           test="LDSC") %>%
    select(VARIABLE, BETA, SE, P, if.sig, test)
  d <- rbind(df.ldsc, df.magma) %>%
    left_join(tmp, by="VARIABLE") %>%
    group_by(VARIABLE) %>%
    mutate(sig=ifelse(min(if.sig)==1, "red","black")) %>%
    ungroup() 
  a <- d %>% distinct(VARIABLE,order, sig) %>% arrange(order) %>% select(sig)
  p1 <- ggplot(d,aes(x=reorder(VARIABLE, order), y=-log10(P),fill=test)) + 
    geom_bar(stat="identity", width=.7, position = position_dodge(width = 0.6)) + 
    coord_flip() + 
    geom_hline(yintercept =-log10(0.05/nrow(df.magma))) + xlab("") +
    ylab(expression('-log'[10]*'(pvalue)')) +
    labs(title="Enrichment of heritability per tissue/cell type")+
    theme(axis.text.y = element_text(color=a$sig)) + 
    scale_fill_manual(values=c("#E69F00", "#0072B2"))
  return(p1) 
}
```
  
Type in paths, trait, tissue etc.  
  
## GTEx_v8  
```{r}
inpath_ldsc <- "/Users/yaoshu/Proj/_Other/Hearing_loss2020/results/03pLDSC/GTEx_v8/tissue_pvalues.txt"
inpath_magma <- "/Users/yaoshu/Proj/_Other/Hearing_loss2020/results/03MAGMA/GTEx_v8/hl2020.gsa.out"
expdataset <- "GTEx_v8"
trait <- "hl2020"
outpath <- "/Users/yaoshu/Proj/_Other/Hearing_loss2020/results/04Output/"
outname_df <- paste(outpath, expdataset, "_", trait, "_results_dataframe.tsv", sep="")
outname_graph <- paste(outpath, expdataset,"_",  trait, "_results_graph.pdf", sep="")

e <- myfun.data(inpath_ldsc, inpath_magma, trait, outname_df)
p <- myfun.plot(e, expdataset)[[2]]
ggsave(p,file=outname_graph)
p
```
  
## Zeisel  
```{r}
inpath_ldsc <- "/Users/yaoshu/Proj/_Other/Hearing_loss2020/results/03pLDSC/Zeisel/tissue_pvalues.txt"
inpath_magma <- "/Users/yaoshu/Proj/_Other/Hearing_loss2020/results/03MAGMA/Zeisel/hl2020.gsa.out"
expdataset <- "Zeisel"
trait <- "hl2020"
outpath <- "/Users/yaoshu/Proj/_Other/Hearing_loss2020/results/04Output/"
outname_df <- paste(outpath, expdataset, "_", trait, "_results_dataframe.tsv", sep="")
outname_graph <- paste(outpath, expdataset,"_",  trait, "_results_graph.pdf", sep="")

e <- myfun.data(inpath_ldsc, inpath_magma, trait, outname_df)
p <- myfun.plot(e, expdataset)[[2]]
ggsave(p,file=outname_graph)
p
```
  
## Zeisel_lvl5  
```{r}
inpath_ldsc <- "/Users/yaoshu/Proj/_Other/Hearing_loss2020/results/03pLDSC/Zeisel_lvl5/tissue_pvalues.txt"
inpath_magma <- "/Users/yaoshu/Proj/_Other/Hearing_loss2020/results/03MAGMA/Zeisel_lvl5/hl2020.gsa.out"
expdataset <- "Zeisel_lvl5"
trait <- "hl2020"
outpath <- "/Users/yaoshu/Proj/_Other/Hearing_loss2020/results/04Output/"
outname_df <- paste(outpath, expdataset, "_", trait, "_results_dataframe.tsv", sep="")
outname_graph <- paste(outpath, expdataset,"_",  trait, "_results_graph.pdf", sep="")

e <- myfun.data(inpath_ldsc, inpath_magma, trait, outname_df)
p <- myfun.plot(e, expdataset)[[2]]
ggsave(p,file=outname_graph)
p

```
  
## HCL  
```{r}
inpath_ldsc <- "/Users/yaoshu/Proj/_Other/Hearing_loss2020/results/03pLDSC/HCL/tissue_pvalues.txt"
inpath_magma <- "/Users/yaoshu/Proj/_Other/Hearing_loss2020/results/03MAGMA/HCL/hl2020.gsa.out"
expdataset <- "HCL"
trait <- "hl2020"
outpath <- "/Users/yaoshu/Proj/_Other/Hearing_loss2020/results/04Output/"
outname_df <- paste(outpath, expdataset, "_", trait, "_results_dataframe.tsv", sep="")
outname_graph <- paste(outpath, expdataset,"_",  trait, "_results_graph.pdf", sep="")

e <- myfun.data(inpath_ldsc, inpath_magma, trait, outname_df)
p <- myfun.plot(e, expdataset)[[2]]
ggsave(p,file=outname_graph)
p

```
  
Plot only significant bars:  
```{r}
f <- myfun.plot(e, expdataset)[[1]] %>% filter(Significant!="None")
p_xlab <- paste("Cell types in ", expdataset, " dataset", sep="")
p1 <- ggplot(f,aes(Tissue,minuslog10P_mean,fill=Significant)) + 
  geom_bar(stat="identity") + coord_flip() + facet_wrap(~Trait) + 
  theme_bw() + geom_hline(yintercept =-log10(0.05/nrow(f))) + xlab("") + 
  ylab(expression('mean -log'[10]*'(pvalue)')) + xlab(p_xlab) + 
  scale_fill_manual(values=c("#E69F00", "#0072B2", "#56B4E9"))
p1
outname_graph2 <- paste(outpath, expdataset,"_",  trait, "_results_graph_onlySignificant.pdf", sep="")
#ggsave(p1,file=outname_graph2)
```
  
## ear_snRNAseq  
```{r}
inpath_ldsc <- "/Users/yaoshu/Proj/_Other/Hearing_loss2020/results/03pLDSC/ear_snRNAseq/tissue_pvalues.txt"
inpath_magma <- "/Users/yaoshu/Proj/_Other/Hearing_loss2020/results/03MAGMA/ear_snRNAseq/hl2020.gsa.out"
expdataset <- "ear_snRNAseq"
trait <- "hl2020"
outpath <- "/Users/yaoshu/Proj/_Other/Hearing_loss2020/results/04Output/"
outname_df <- paste(outpath, expdataset, "_", trait, "_results_dataframe.tsv", sep="")
outname_graph <- paste(outpath, expdataset,"_",  trait, "_results_graph.pdf", sep="")

e <- myfun.data(inpath_ldsc, inpath_magma, trait, outname_df)
p <- myfun.plot(e, expdataset)[[2]]
ggsave(p,file=outname_graph)
p

```
  
```{r}
inpath_magma <- "/Users/yaoshu/Proj/_Other/Hearing_loss2020/results/03MAGMA/ear_snRNAseq/hl2020.gsa.out"
inpath_ldsc <- "/Users/yaoshu/Proj/_Other/Hearing_loss2020/results/03pLDSC/ear_snRNAseq/tissue_pvalues.txt"
plot_out <- "/Users/yaoshu/Proj/_Other/Hearing_loss2020/results/04Output/"
p1 <- plot.magma.ldsc(inpath_magma, inpath_ldsc, plot_out, "ear_snRNAseq")
p1
```
  
```{r}
inpath_magma <- "/Users/yaoshu/Proj/_Other/Hearing_loss2020/results/03MAGMA/ear_snRNAseq/hl2020.gsa.out"
inpath_ldsc <- "/Users/yaoshu/Proj/_Other/Hearing_loss2020/results/03pLDSC/ear_snRNAseq_35up_10down/tissue_pvalues.txt"
plot_out <- "/Users/yaoshu/Proj/_Other/Hearing_loss2020/results/04Output/"
p1 <- plot.magma.ldsc(inpath_magma, inpath_ldsc)
p1
outfile <- paste(plot_out,"ear_snRNAseq_35up_10down","_p1.MAGMA_LDSC.pdf", sep="")
ggsave(p1, file=outfile,
         width=6, height=6)
```
